---
title: "Stroke Prediction"
author: "Trevor Okinda"
date: "2024"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **Project Name**                             | Stroke Prediction |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

### Source: 

The dataset that was used can be downloaded here: *\<https://www.kaggle.com/datasets/fedesoriano/stroke-prediction-dataset\>*

### Reference:

*\<fedesoriano. (2020). Stroke Prediction Dataset. Retrieved from Kaggle https://www.kaggle.com/datasets/fedesoriano/stroke-prediction-dataset\>\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

# Exploratory Data Analysis
## Load dataset
```{r Load dataset}
# Load dataset
stroke_data <- read.csv("stroke_data.csv", colClasses = c(
  gender = "factor",
  age = "numeric",
  hypertension = "factor",
  heart_disease = "factor",
  ever_married = "factor",
  work_type = "factor",
  Residence_type = "factor",
  avg_glucose_level = "numeric",
  bmi = "numeric",
  smoking_status = "factor",
  stroke = "factor"
))

# Display the structure of the dataset
str(stroke_data)

# View the first few rows of the dataset
head(stroke_data)

# View the dataset in a separate viewer window
View(stroke_data)
```

## Measures of Frequency
```{r MOF}
# Measures of Frequency
# Count of stroke events
stroke_count <- table(stroke_data$stroke)
print("Frequency of stroke events:")
print(stroke_count)
```

## Measures of Central Tendency
```{r MOCT}
# Measures of Central Tendency
# Mean age
mean_age <- mean(stroke_data$age, na.rm = TRUE)
print("Mean age:")
print(mean_age)

# Median avg_glucose_level
median_glucose <- median(stroke_data$avg_glucose_level, na.rm = TRUE)
print("Median avg_glucose_level:")
print(median_glucose)
```

## Measures of Distribution
```{r MOD}
# Measures of Distribution
# Standard deviation of BMI
sd_bmi <- sd(stroke_data$bmi, na.rm = TRUE)
print("Standard deviation of BMI:")
print(sd_bmi)

# Range of avg_glucose_level
range_glucose <- range(stroke_data$avg_glucose_level, na.rm = TRUE)
print("Range of avg_glucose_level:")
print(range_glucose)
```

## Measures of Relationship
```{r MOR}
# Measures of Relationship
# Correlation between age and avg_glucose_level
correlation_age_glucose <- cor(stroke_data$age, stroke_data$avg_glucose_level, use = "complete.obs")
print("Correlation between age and avg_glucose_level:")
print(correlation_age_glucose)

# Association between smoking status and stroke using Chi-square test
chi_square_smoking_stroke <- chisq.test(table(stroke_data$smoking_status, stroke_data$stroke))
print("Chi-square test for association between smoking status and stroke:")
print(chi_square_smoking_stroke)
```

## ANOVA
```{r ANOVA}
# Check the levels of work_type
levels(stroke_data$work_type)

# Perform ANOVA
anova_result <- aov(avg_glucose_level ~ work_type, data = stroke_data)
print("ANOVA results:")
print(summary(anova_result))
```

## Plots
```{r Plots}
# Load required libraries
library(ggplot2)
library(GGally)

# Univariate Plots
# Histogram of age
ggplot(stroke_data, aes(x = age)) +
  geom_histogram(fill = "skyblue", color = "black", bins = 30) +
  labs(title = "Histogram of Age",
       x = "Age",
       y = "Frequency")

# Boxplot of avg_glucose_level
ggplot(stroke_data, aes(x = 1, y = avg_glucose_level, fill = stroke)) +
  geom_boxplot() +
  labs(title = "Boxplot of Average Glucose Level by Stroke Status",
       x = NULL,
       y = "Average Glucose Level",
       fill = "Stroke Status") +
  theme(axis.text.x = element_blank())  # Hide x-axis label

# Multivariate Plot
# Scatterplot of age vs. avg_glucose_level colored by stroke status
ggplot(stroke_data, aes(x = age, y = avg_glucose_level, color = stroke)) +
  geom_point() +
  labs(title = "Scatterplot of Age vs. Average Glucose Level by Stroke Status",
       x = "Age",
       y = "Average Glucose Level",
       color = "Stroke Status")

# Pairwise scatterplot matrix of numerical variables
num_vars <- c("age", "avg_glucose_level", "bmi")
# Pairwise scatterplot matrix of numerical variables including 'stroke'
ggpairs(stroke_data, columns = num_vars, mapping = aes(color = stroke))
```

# Preprocessing and Data Transformation
## Missing Values
```{r Missing Values}
# Check for missing values in each column
missing_values <- sapply(stroke_data, function(x) sum(is.na(x)))
print("Missing values in each column:")
print(missing_values)

# Total missing values in the dataset
total_missing <- sum(missing_values)
print("Total missing values in the dataset:")
print(total_missing)
```

## Imputation
```{r Imputation}
# Impute missing values in 'bmi' column with mean
mean_bmi <- mean(stroke_data$bmi, na.rm = TRUE)
stroke_data$bmi[is.na(stroke_data$bmi)] <- mean_bmi

# Confirm that missing values have been imputed
missing_values_after_imputation <- sum(is.na(stroke_data$bmi))
print("Missing values in 'bmi' column after imputation:")
print(missing_values_after_imputation)
```

